---
title: "newton_raphson"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{newton_raphson}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(scipen=999)
```

```{r setup}
suppressPackageStartupMessages({
  library(ipfr)
  library(dplyr)
})
```

# ipu_nr

The function `ipu_nr` only differs from `ipu` in the method used to balance
primary and secondary targets. In addition, because it was created merely for
comparison to `ipu`, it has not been maintained (and is now an internal package
function). Note in the ASU example creation code that fields `geo_region`
and `pid` are required, which is no longer true for `ipu`.

The following code block re-creates the 
[Arizona](http://www.scag.ca.gov/Documents/PopulationSynthesizerPaper_TRB.pdf)
example.

```{r}
hh_seed <- tibble(
  geo_region = 1,
  pid = c(1:8),
  hhtype = c(1, 1, 1, 2, 2, 2, 2, 2)
)

per_seed <- tibble(
  pid = c(1, 1, 1, 2, 2, 3, 3, 3, 4, 4, 4, 5, 5, 5, 6, 6, 7, 7, 7, 7, 7, 8, 8),
  pertype = c(1, 2, 3, 1, 3, 1, 1, 2, 1, 3, 3, 2, 2, 3, 1, 2, 1, 1, 2, 3, 3, 1, 2)
)

hh_targets <- list()
hh_targets$hhtype <- tibble(
  geo_region = 1,
  `1` = 35,
  `2` = 65
)

per_targets <- list()
per_targets$pertype <- tibble(
  geo_region = 1,
  `1` = 91,
  `2` = 65,
  `3` = 104
)
```

As in the example in the `common_ipf_problems` vignette, we modify the Arizona
example (which is balanced) to double the person targets. This creates a
significant imbalance that standard approaches struggle with.

```{r}
new_per_targets <- per_targets
new_per_targets$pertype <- per_targets$pertype %>%
  mutate_at(
    .vars = vars(`1`, `2`, `3`),
    .funs = list(~. * 2)
  )
```

While `ipu` balances the secondary targets directly using `secondary_importance`,
`ipu_nr` uses an iterative approach and the `target_priority` argument. 

By default, all target tables have an equally high priority, which means that the
algorithm will attempt to match all targets exactly. However, `target_priority`
can be modified in several ways. In the code below, a data frame is used to
assign the `hhtype` target a higher priority. (If using a data frame, the column
names must be `target` and `priority`.) A simple named list can also be used
(both options shown below).

```{r}
# Option 1: a data frame
target_priority <- tibble(
  target = c("hhtype", "pertype"),
  priority = c(10000, 10)
)

# Options 2: use a named list
target_priority <- list()
target_priority$hhtype <- 10000
target_priority$pertype <- 10

result <- ipfr:::ipu_nr(
  hh_seed, hh_targets, per_seed, new_per_targets, max_iterations = 30,
  target_priority = target_priority
)
```

As `ipu_nr` runs, it relaxes the target constraints on `pertype` much faster
than on `hhtype`. As a result, the final weights will match the household type
much closer. The two methods generally match targets to the same degree, but
often lead to very different distributions of weight ratios. In addition, `ipu`
tends to reach convergence levels around .1 %RMSE faster than `ipu_nr`, but for
levels below that, `ipu_nr` tends to be faster. For IPF, convergence below .1
%RMSE is of questionable utility.

```{r}
result
```
