---
title: "Using ipfr"
author: "Kyle Ward"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ipfr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE,echo=TRUE,
                      message=FALSE,warning=FALSE,error=FALSE)
options(scipen=999) # removes sci notation
```

## Introduction

This package provides a generic implimentation of the iterative proportional 
fitting algorithm or [IPF](https://en.wikipedia.org/wiki/Iterative_proportional_fitting).

## Example data creation

This section creates a random seed table and fixed marginal values to illustrate 
how the package is used.

### Seed table creation
The seed table is the starting point for the IPF procedure.  In this example, 
we will use a made up summary table of survey data.  This fake table is a count 
of households by the number persons, workers and vehicles.
```{r "seed creation"}
library(dplyr)
library(ipfr)
library(htmlTable)

seed <- expand.grid(
  siz = c(1, 2, 3, 4),
  wrk = c(0, 1, 2, 3),
  veh = c(0, 1, 2, 3)
) %>% tbl_df()

set.seed(1)
seed$count <- sample(1:10, nrow(seed), replace = TRUE)

seed %>%
  head() %>%
  htmlTable(
  align = "cccr", col.rgroup = c("none", "#F7F7F7"), rnames = FALSE
)
```


### Marginal creation
The number of households by size (e.g., 1-person, 2-person, etc.) is referred to
as a marginal distribution.  Often, from the Census, we know the total number of 
households by each individual marginal.  For example, we know the number of 
households by size:

  * 1 person: 150 households
  * 2 persons: 150 households
  * 3 persons: 100 households
  * 4+ persons: 100 households

This information is created below. For the distribution of households by size, 
number of workers, and number of vehicles.  Importanlty, the "marginal" column 
is a concatenation of the marginal name in the seed table ("siz", "wrk", and 
"veh") and the numbers describing the categories within that marginal found in 
the seed table (e.g., 0, 1, 2, etc.)
```{r "marginal creation"}

marginals <- data_frame(
  marginal = c(
    "siz1", "siz2", "siz3", "siz4",
    "wrk0", "wrk1", "wrk2", "wrk3",
    "veh0", "veh1", "veh2", "veh3"
    ),
  value = c(
    100, 100, 150, 150,
    100, 150, 150, 100,
    75, 175, 150, 100
  )
)

marginals %>% htmlTable(
  align = "cr", col.rgroup = c("none", "#F7F7F7"), rnames = FALSE
)
```

## Using ipfr
We know the count of each marginal individually, but assume we don't have 
official information from the Census about the joint distribution.  What we want
to do is use the joint distribution from our survey, but ensure it agrees with 
the marginal information from the Census.

An initial test comparing size shows they do not agree.
```{r "initial test"}
siz <- marginals %>%
    mutate(
      category = as.numeric(gsub("[A-z]", "", marginal)),
      marginal = gsub("[0-9]", "", marginal)
    ) %>%
  filter(marginal == "siz") %>%
  select(siz = category, value)
  

seed %>%
  select(siz, count) %>%
  group_by(siz) %>%
  summarize(total = sum(count)) %>%
  left_join(siz) %>%
  rename(survey = total, census = value) %>%
  htmlTable(
    align = "crr", col.rgroup = c("none", "#F7F7F7"), rnames = FALSE
  )
```

We can use the ipfr package to change this.
```{r}
seed$NewWeight <- ipf(seed, weight_var = "count", marginals)
```

Once complete, we can summarize the final table to see if it worked.
```{r}
seed %>%
  group_by(siz) %>%
  summarize(total = sum(NewWeight)) %>%
  left_join(siz) %>%
  rename(survey = total, census = value) %>%
  mutate(survey = round(survey, 3)) %>%
  htmlTable(
    align = "crr", col.rgroup = c("none", "#F7F7F7"), rnames = FALSE
  )
```

Not only does the marginal distribution of persons match, but also workers and
vehicles.
```{r, echo=FALSE}
wrk <- marginals %>%
    mutate(
      category = as.numeric(gsub("[A-z]", "", marginal)),
      marginal = gsub("[0-9]", "", marginal)
    ) %>%
  filter(marginal == "wrk") %>%
  select(wrk = category, value)

seed %>%
  group_by(wrk) %>%
  summarize(total = sum(NewWeight)) %>%
  left_join(wrk) %>%
  rename(survey = total, census = value) %>%
  mutate(survey = round(survey, 3)) %>%
  htmlTable(
    align = "crr", col.rgroup = c("none", "#F7F7F7"), rnames = FALSE
  )

veh <- marginals %>%
    mutate(
      category = as.numeric(gsub("[A-z]", "", marginal)),
      marginal = gsub("[0-9]", "", marginal)
    ) %>%
  filter(marginal == "veh") %>%
  select(veh = category, value)

seed %>%
  group_by(veh) %>%
  summarize(total = sum(NewWeight)) %>%
  left_join(veh) %>%
  rename(survey = total, census = value) %>%
  mutate(survey = round(survey, 3)) %>%
  htmlTable(
    align = "crr", col.rgroup = c("none", "#F7F7F7"), rnames = FALSE
  )
```

## Addressing common IPF problems

### Zero weights
IPF works by successively updating the table weights by factors.  Cells with a 
zero weight cannot be modified by this process.  As the number of zero weights 
increase, the flexibility of the process is reduced, and convergence becomes 
more difficult. `ipfr` solves this problem by setting a minimum weight for all 
cells to `.0001`.  This minimum weight can be adjusted using the `min_weight` 
parameter and should be arbitrarily small compared to your seed table weights.

### Marginal agreement
In this vignette, the marginal distributions all added up to 500 households.
If they disagreed on the total number of households, the standard process could 
not converge.

Occasionally, marginal distributions might disagree on the totals, particularly 
if they are derived from different sources.  When that happens, this function 
will still converge by matching the total of the first marginal table supplied 
while using the percentage distribution from the remaining tables.  A warning 
will be displayed when this happens.

```{r, warning=TRUE}
# Increase 1-person households from 100 to 500
persons <- data_frame(
  persons = c(1, 2, 3, 4),
  value = c(500, 100, 150, 150)
)

marginals <- list(
  persons = persons,
  workers = workers,
  vehicles = vehicles
)

seed$NewWeight <- ipf(df, "count", marginals)
```

```{r, echo=FALSE}
seed %>%
  group_by(persons) %>%
  summarize(total = sum(NewWeight)) %>%
  left_join(persons) %>%
  rename(survey = total, census = value) %>%
  mutate(survey = round(survey, 3)) %>%
  htmlTable(
    align = "crr", col.rgroup = c("none", "#F7F7F7"), rnames = FALSE
  )

seed %>%
  group_by(workers) %>%
  summarize(total = sum(NewWeight)) %>%
  left_join(workers) %>%
  rename(survey = total, census = value) %>%
  mutate(survey = round(survey, 3)) %>%
  htmlTable(
    align = "crr", col.rgroup = c("none", "#F7F7F7"), rnames = FALSE
  )

seed %>%
  group_by(vehicles) %>%
  summarize(total = sum(NewWeight)) %>%
  left_join(vehicles) %>%
  rename(survey = total, census = value) %>%
  mutate(survey = round(survey, 3)) %>%
  htmlTable(
    align = "crr", col.rgroup = c("none", "#F7F7F7"), rnames = FALSE
  )
```