---
title: "Using ipf_multi"
author: "Kyle Ward"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ipf-multi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE,echo=TRUE,
                      message=FALSE,warning=FALSE,error=FALSE)
options(scipen=999) # removes sci notation
```

## Introduction
`ipf_multi` is a wrapper to the `ipf` function included to support specific applications.  For example, in travel modeling, IPF is often performed on each zone in the model to create cross-classified distributions of households.

## Example data creation

### Seed table creation
The seed table provides an observed, usually regional, three-way distribution of households by size, workers, and vehicles.  It uses the same format as `ipf`.
```{r "seed creation"}
library(dplyr)
library(ipfr)
library(htmlTable)

seed <- expand.grid(
  siz = c(1, 2),
  wrk = c(0, 1),
  veh = c(0, 1)
) %>% tbl_df()

set.seed(1)
seed$count <- sample(1:10, nrow(seed), replace = TRUE)

seed %>%
  head() %>%
  htmlTable(
  align = "cccr", col.rgroup = c("none", "#F7F7F7"), rnames = FALSE
)
```

### Marginal table creation
For `ipf_multi`, the marginal table looks different from the marginal table in `ipf`.  Each row now represents a zone where IPF will be applied.  The marginals are spread across the columns.  In this example, there are two zones: 1000 and 1001.
```{r}
margTbl <- data_frame(
  ID = c(1000, 1001),
  siz1 = c(100, 150),
  siz2 = c(150, 100),
  wrk0 = c(50, 200),
  wrk1 = c(200, 50),
  veh0 = c(75, 175),
  veh1 = c(175, 75)
)

margTbl %>%
  htmlTable(
    col.rgroup = c("none", "#F7F7F7"), rnames = FALSE
  )
```

## Using ipf_multi
These two sources of information can be fed into `ipf_multi`.  `weight_var` names the field in the seed table with the count or percentage distribution.  `id_field` names the ID field in the marginal table.  A sample of the result is shown below.
```{r}
tbl <- ipf_multi(seed, weight_var = "count", margTbl, id_field = "ID")

tbl %>% 
  mutate_each(funs(round(., digits = 2))) %>%
  select(1:4) %>%
  htmlTable(
  col.rgroup = c("none", "#F7F7F7"), rnames = FALSE
)
```
Because `ipf_multi` calls `ipf`, all the same error checks and warnings are performed for each row in the marginal table.  Refer to the `ipf` vignette for more information on what those are.
