---
title: "Using ipf"
author: "Kyle Ward"
date: "`r Sys.Date()`"
output: 
  knitr:::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{ipf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE,echo=TRUE,
                      message=TRUE,warning=TRUE,error=TRUE)
options(scipen=999) # removes sci notation
```

# Introduction

This package provides a generic implimentation of the iterative proportional 
fitting algorithm or
[IPF](https://en.wikipedia.org/wiki/Iterative_proportional_fitting) in the
`ipf()` function. It also provides an iterative proportional updating algorithm
based on on the paper from Arizona State University 
([IPU](http://www.scag.ca.gov/Documents/PopulationSynthesizerPaper_TRB.pdf)) for
balancing household- and person-level marginals in the `ipu()` function.

```{r, message=FALSE,warning=FALSE,error=FALSE}
library(dplyr)
library(tidyr)
library(ipfr)
library(htmlTable)
```

# IPU

Iterative proportional updating is a method developed by Arizona State University
that allows the IPF procedure to match household- and person-level marginals. In
the basic IPF procedure, all marginal distributions must describe the same thing
(e.g. households). IPU allows you to say, for example, that a zone needs a total
household count of 500, but also needs 800 people.

## Example 1: Simple Example

This example creates a random seed table and target values to illustrate how the
package is used. The targets are specified for two separate geographies
(`geo_clusters`). Any field name can be used as long as it:

  * starts with "geo_"
  * Is included in both the target table(s) and seed table

This simple example only has one target marginal distribution, and could be
solved directly without `ipu`. However, it is designed to show the basics needed
to run the function.

### Seed table creation
The seed table is the starting point for the IPF procedure. In this example, we
make up some survey data.

  * Each row represents a household
  * The household ID column is named/renamed to `pid` ("primary ID")
  * The geography field is inlcluded and starts with "geo_" (`geo_cluster`)

```{r}
hh_seed <- data_frame(
  pid = c(1, 2, 3, 4),
  siz = c(1, 2, 1, 1),
  weight = c(1, 1, 1, 1),
  geo_cluster = c(1, 1, 2, 2)
)

hh_seed %>%
  htmlTable(
    rnames = FALSE, caption = "Seed Table with Cluster Appended"
  )
```

### Target creation
The number of households by size (e.g., 1-person, 2-person, etc.) is referred to
as a marginal distribution.  Often, from the Census, we know the total number of 
households by each individual marginal.  For example, we know the number of 
households by size:

  * 1 person households 
    * 75 in geographic cluster 1
    * 100 in cluster 2
  * 2 personshouseholds
    * 25 in cluster 1
    * 150 in cluster 2

This information becomes the target that the IPU process tries to match.

Marginal targets are specified below for each cluster:

  * The geography field `geo_cluster` matches the seed table
  * The name of the table in the list links it to the `siz` column of the seed
  * The column names are the values that show up in the seed's `siz` column

```{r}
hh_targets <- list()
hh_targets$siz <- data_frame(
  geo_cluster = c(1, 2),
  `1` = c(75, 100),
  `2` = c(25, 150)
)

hh_targets$siz %>%
  htmlTable(
    rnames = FALSE, caption = "Target Table"
  )
```

### Run IPU

Run `ipu()`. Note that it throws an error. When providing cluster values in
the seed table, `ipfr` checks to make sure that every marginal category appears
in every cluster. In this case, the seed table does not have a 2-person household
in cluster 2. This makes it impossible to match that marginal target.

```{r}
result <- ipu(hh_seed, hh_targets)
```

Fix the seed table and re-run `ipu`. The results match the targets we provided.

```{r}
hh_seed <- data_frame(
  pid = c(1, 2, 3, 4),
  siz = c(1, 2, 2, 1),
  weight = c(1, 1, 1, 1),
  geo_cluster = c(1, 1, 2, 2)
)

result <- ipu(hh_seed, hh_targets)
```

`ipu()` returns a named list.

```{r}
names(result)
```

The first element is the resulting weight table. The primary seed has three
columns appended:

  * weight
    * The expanded weight of the record.
  * avg_weight
    * The average weight for the geography (total target / number of seed records)
  * weight_factor
    * The weight divided by the average weight.

```{r}
result$weight_tbl %>%
  htmlTable(
    rnames = FALSE
  )
```

The second element is a histogram of the `weight_factor`. This provides a quick
overview of the distribution of weights.

```{r}
result$weight_dist
```

The next element is a comparison back to the targets provided. With complex
seed and target tables, this makes investigating results quick and easy.

```{r}
result$primary_comp %>%
  htmlTable(
    rnames = FALSE
  )
```

If secondary targets are provided to `ipu()`, a fourth item in the list will
contain a `secondary_comp` table.

## Example 2: The Arizona Paper Example

In household survey expansion, it is common to want to control for certain 
features that describe households, (like size), while controlling for other 
attributes that describe people (like age). This is possible with the `ipu()`
function.

This example is taken directly from the Arizona paper on page 20:
http://www.scag.ca.gov/Documents/PopulationSynthesizerPaper_TRB.pdf

In this example, household type could represent size (e.g. 1-person and 2-person
households). Person type could represent age groups (e.g. under 18, between
18 and 50, and over 50).

The code block below re-creates the seed and target tables for both persons
and households.

  * Only a single geography is used (`geo_region`)
  * Both seed tables have the `pid` field
    * The `pid` field in the persons seed table links to the household seed

```{r example 1 inputs}
hh_seed <- data_frame(
  geo_region = 1,
  pid = c(1:8),
  hhtype = c(1, 1, 1, 2, 2, 2, 2, 2)
)

per_seed <- data_frame(
  pid = c(1, 1, 1, 2, 2, 3, 3, 3, 4, 4, 4, 5, 5, 5, 6, 6, 7, 7, 7, 7, 7, 8, 8),
  pertype = c(1, 2, 3, 1, 3, 1, 1, 2, 1, 3, 3, 2, 2, 3, 1, 2, 1, 1, 2, 3, 3, 1, 2)
)

hh_targets <- list()
hh_targets$hhtype <- data_frame(
  geo_region = 1,
  `1` = 35,
  `2` = 65
)

per_targets <- list()
per_targets$pertype <- data_frame(
  geo_region = 1,
  `1` = 91,
  `2` = 65,
  `3` = 104
)
```

In the interst of keeping vignette build time short, the `ipu()` algorithm is 
only run for 30 iterations. After running for 400 or more iterations, the 
results match closely to those shown in the paper.

  * The household seed table is the `primary_seed`
  * The household target list is the `primary_target`
  * The person seed table is the `secondary_seed`
  * The person target list is the `secondary_target`

```{r example 1 ipu}
result <- ipu(hh_seed, hh_targets, per_seed, per_targets, max_iterations = 30)
```

The first table shows the result. The second table shows the primary comparison
table. Since we added secondary seeds and targets, the output now contains
a secondary comparison table. Feel free to run the code chunk above for
400 or more iterations and then look again.

```{r example 1 results}
result$weight_tbl %>%
  mutate(weight = round(weight, 2)) %>%
  htmlTable(
    rnames = FALSE, caption = "Result"
  )
  
result$primary_comp %>% 
  mutate(result = round(result, 2)) %>%
  htmlTable(
    rnames = FALSE, caption = "Primary Comparison"
  )

result$secondary_comp %>% 
  mutate(result = round(result, 2)) %>%
  htmlTable(
    rnames = FALSE, caption = "Secondary Comparison"
  )
```


## Example 3: Using Multiple Geographies

`ipu()` allows different geographies to be specified for different marginal
tables. There are a few rules that make this possible, but in short, the geo
field on each target table tells the algorithm which scale to constrain to.

All of the following rules are checked by the algorithm a warning message will
show if one is violated.

  * The primary/household seed table must contain all geo fields used by any
    target table and the `pid` field.
    * Do **not** put any geo fields on the secondary/person seed table.
    * This prevents potential errors/inconsistencies between seed tables.
  * All fields that designate geographies must start with "geo_" e.g.
    * geo_cluster
    * geo_region
    * geo_state
  * Each target table must have a geo field that is present in the primary seed 
    table

To demonstrate, the Arizona example from example 1 is modified to add two
different clusters for household controls but to still control the person
targets at the regional level.

```{r example 2 inputs}
# Modifying example 1 for example 2

# Repeat the hh_seed to create cluster 1 and 2 households
hh_seed <- hh_seed %>%
  rename(geo_cluster = geo_region)
hh_seed <- bind_rows(
  hh_seed,
  hh_seed %>% 
    mutate(geo_cluster = 2, pid = pid + 8)
)
hh_seed$geo_region = 1

hh_seed %>%
  htmlTable(
    rnames = FALSE, caption = "Modified household seed table"
  )

# Repeat the household targets for two clusters
hh_targets$hhtype <- bind_rows(hh_targets$hhtype, hh_targets$hhtype)
hh_targets$hhtype <- hh_targets$hhtype %>%
  rename(geo_cluster = geo_region) %>%
  mutate(geo_cluster = c(1, 2))

hh_targets$hhtype %>%
  htmlTable(
    rnames = FALSE, caption = "Modified household targets"
  )

# Repeat the per_seed to create cluster 1 and 2 persons
per_seed <- bind_rows(
  per_seed,
  per_seed %>% 
    mutate(pid = pid + 8)
)

per_seed %>%
  head() %>%
  htmlTable(
    rnames = FALSE, caption = "Modified person seed (first few rows)"
  )

# Double the regional person targets
per_targets$pertype <- per_targets$pertype %>%
  mutate_at(
    .vars = vars("1", "2", "3"),
    .funs = funs(. * 2)
  )

per_targets$pertype %>%
  htmlTable(
    rnames = FALSE, caption = "Modified person targets"
  )
```

Run the IPU algorithm. Again, for vignette build time, only 30 iterations are
performed. Run the code yourself with `max_iterations` set to 600 to see the
converged result.

```{r example 2 ipu}
result <- ipu(hh_seed, hh_targets, per_seed, per_targets, max_iterations = 30)
```

The tables below show the results compared back to targets. More iterations
would make a better match.

```{r example 2 results}
result$primary_comp %>%
  mutate(result = round(result, 2)) %>%
  htmlTable(
    rnames = FALSE, caption = "Primary Comparison"
  )

result$secondary_comp %>%
  mutate(result = round(result, 2)) %>%
  htmlTable(
    rnames = FALSE, caption = "Secondary Comparison"
  )
```

# How IPU addresses common IPF problems

This section will show how `ipu()` addresses some common problems found in basic
ipf procedures. It uses the example data from the first example.

## Zero weights

IPF works by successively multiplying the table weights by factors.  Cells with 
a zero weight cannot be modified by this process.  As the number of zero weights 
increase, the flexibility of the process is reduced, and convergence becomes 
more difficult. `ipfr` solves this problem by setting a minimum weight for all 
cells to `.0001`.  This minimum weight can be adjusted using the `min_weight` 
parameter and should be arbitrarily small compared to your seed table weights.

## Missing seed information

Not every combination of marginal categories is required to be included in the 
seed table; however, at least one observation of each category must exist.  For example, the combination:

  * siz = 1
  * wrk = 1
  * veh = 0

may not have been observed in the survey, and thus may be missing from the seed 
table.  As long as other combinations of size-1 households exist (e.g. with 0 workers and 1 vehicle), `ipfr` will work fine.  On the other hand, if there are no observations of any size-1 households, `ipfr` will stop with an error message.

See the first IPU example to see how it works.

## Marginal agreement

In the above examples, are marginals disagreed on totals (either persons or 
households). If they disagreed on the total, the standard IPF process could not
converge.

Occasionally, marginal distributions might disagree slightly on the totals, 
particularly if they are derived from different sources.  When that happens, 
`ipu()` will still converge by scaling all tables to match the total of the
first table. This eliminates the painful task of ensuring all marginals match
exactly.

In the example below, the size marginal sums to a total of 100 households.
The vehicle marginal sums to 300. With the `verbose` option set to `TRUE`,
a message will be displayed telling which, if any, target tables are scaled.

```{r, warning=TRUE}
hh_seed <- data_frame(
  geo_region = 1,
  pid = c(1:8),
  hhsiz = c(1, 1, 1, 2, 2, 2, 2, 2),
  hhveh = c(0, 2, 1, 1, 1, 2, 1, 0)
)

hh_targets <- list()
hh_targets$hhsiz <- data_frame(
  geo_region = 1,
  `1` = 35,
  `2` = 65
)
hh_targets$hhveh <- data_frame(
  geo_region = 1,
  `0` = 100,
  `1` = 100,
  `2` = 100
)

result <- ipu(hh_seed, hh_targets, max_iterations = 30, verbose = TRUE)
```

Importantly, the performance measures below compare the result to the scaled
target not the original. Note that the vehicle targets have been scaled down.

```{r}
result$primary_comp %>%
  mutate_at(
    .vars = vars(target, result),
    .funs = funs(round(., 2))
  ) %>%
  htmlTable(
    rnames = FALSE, caption = "Scaled Results"
  )
```

## Extreme Weights

Often, it is preferable to constrain weights so that certain, under-sampled
observations to do not end up with extreme weights. `ipu()` supports this by
using the `min_factor` and `max_factor` variables.

First, the average weight is calculated **per geography** based on the total of 
the target tables divided by the number of records in the seed table. Then, the 
max and min factors set a cap and floor based on a multiple of that average.

Common values to use are:

  * max_factor = 5 (5x the average weight)
  * min_factor = .2 (1/5 the average weight)

However, care should be taken when moving these variables from their default
values. These variables impose another constraint on the algorithm and increase
the chance of failure. In the example below, very strict values are used with
the same seed and target data from IPU Example 1.

Values of 1.2 and .8 mean that all weights must be within 20% of the average
weight.

```{r}
hh_seed <- data_frame(
  pid = c(1, 2, 3, 4),
  siz = c(1, 2, 2, 1),
  weight = c(1, 1, 1, 1),
  geo_cluster = c(1, 1, 2, 2)
)

hh_targets <- list()
hh_targets$siz <- data_frame(
  geo_cluster = c(1, 2),
  `1` = c(75, 100),
  `2` = c(25, 150)
)

result <- ipu(hh_seed, hh_targets, max_iterations = 10,
              max_factor = 1.2, min_factor = .8)
```

Consider the effect on geo_cluster 1. With a total target of 100 households and
two records in the seed table, the average weight is 50. This means that the
weights must be between 40 and 60. The algorithm does not have enough flexibility
to meet the controls.

```{r}
result$primary_comp %>%
  htmlTable(
    rnames = FALSE, caption = "Constrained Results"
  )
```

A second problem can arrise from capping weights based on the average weight. In
the example below, I change the targets so that, for geo_cluster 1, they are 
very unbalanced. Cluster 1 now has 100,000 1-person households but only 5 
2-person households. 

```{r}
hh_targets <- list()
hh_targets$siz <- data_frame(
  geo_cluster = c(1, 2),
  `1` = c(100000, 100),
  `2` = c(5, 150)
)

result <- ipu(hh_seed, hh_targets, max_iterations = 10,
              max_factor = 5, min_factor = .2)

result$primary_comp %>%
  htmlTable(
    rnames = FALSE, caption = "Unbalanced Targets"
  )
```

Even with reasonable values for the weight caps, the minimum allowable weight is
much higher than 5. This is an extreme example, and is unlikely to be an issue
in applications related to housing and population - the targets are generally on
the same scale. However, when expanding a through-trip table, it is common to
have some external stations with large targets and others with small. In these
cases, it is advisable to leave the scale arguments at their default values.

# IPF

**Note: `ipf()` is being deprecated in favor of `ipu()`.**

## Example 1: Basics

This section creates a random seed table and target values to illustrate how the
function is used. The targets are specified for three separate clusters, which 
could represent model TAZs, survey districts, or some other grouping of 
geographic areas.

Seed table creation.

```{r seed creation}
seed <- expand.grid(
  siz = c(1, 2, 3, 4),
  wrk = c(0, 1, 2, 3),
  veh = c(0, 1, 2, 3)
) %>% tbl_df()

set.seed(1)
seed$weight <- sample(1:10, nrow(seed), replace = TRUE)

seed %>%
  head() %>%
  htmlTable(
  align = "cccr", col.rgroup = c("none", "#F7F7F7"), rnames = FALSE
)
```

Target creation.

```{r "marginal creation"}

targets <- list()
targets$siz <- data_frame(
  cluster = c(1, 2, 3),
  `1` = c(100, 150, 0),
  `2` = c(100, 100, 0),
  `3` = c(150, 100, 0),
  `4` = c(150, 150, 0)
)
targets$wrk <- data_frame(
  cluster = c(1, 2, 3),
  `0` = c(100, 100, 0),
  `1` = c(150, 100, 0),
  `2` = c(150, 150, 0),
  `3` = c(100, 150, 0)
)
targets$veh <- data_frame(
  cluster = c(1, 2, 3),
  `0` = c(75, 175, 0),
  `1` = c(175, 75, 0),
  `2` = c(150, 150, 0),
  `3` = c(100, 100, 0)
)

targets$siz %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Size Targets"
  )  
```

Run `ipf()`.

```{r}
table <- ipf(seed, targets, verbose = TRUE)
```

Results.

```{r}
table %>%
  mutate_all(funs(round(., 2))) %>%
  head() %>%
  htmlTable(
    align = "crr", rnames = FALSE
  )
```

```{r}
table %>%
  select(-c(wrk, veh)) %>%
  group_by(cluster, siz) %>%
  summarize(total = round(sum(weight), 0)) %>%
  spread(key = siz, value = total) %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Size: Result"
  )

targets$siz %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Size: Target"
  )
```

Not only does the marginal distribution of size match, but also workers and vehicles.
```{r, echo=FALSE}
# Workers
table %>%
  select(-c(siz, veh)) %>%
  group_by(cluster, wrk) %>%
  summarize(total = round(sum(weight), 0)) %>%
  spread(key = wrk, value = total) %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Workers: Result"
  )

targets$wrk %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Workers: Target"
  )

# Vehciles
table %>%
  select(-c(siz, wrk)) %>%
  group_by(cluster, veh) %>%
  summarize(total = round(sum(weight), 0)) %>%
  spread(key = veh, value = total) %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Vehicles: Result"
  )

targets$veh %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Vehciles: Target"
  )
```

```{r}
# Saving seed and target for later examples.
first_seed <- seed
first_targets <- targets
```

## Example 2: Survey Expansion using Geographical Clusters

Seed table creation.

```{r}
seed <- data_frame(
  hhid = c(1, 2, 3, 4),
  siz = c(1, 2, 2, 1),
  weight = c(1, 1, 1, 1),
  cluster = c(1, 1, 2, 2)
)

seed %>%
  htmlTable(
    rnames = FALSE, caption = "Seed Table with Cluster Appended"
  )
```

Targets for each cluster.

```{r}
targets <- list()
targets$siz <- data_frame(
  cluster = c(1, 2),
  `1` = c(75, 100),
  `2` = c(25, 150)
)
```

Run `ipf`

```{r}
result <- ipf(seed = seed, targets = targets)
```


Results.

```{r}
result %>%
  htmlTable(
    rnames = FALSE, caption = "Survey Expansion to Clusters"
  )
```

