---
title: "Using ipf"
author: "Kyle Ward"
date: "`r Sys.Date()`"
output: 
  knitr:::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{ipf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE,echo=TRUE,
                      message=TRUE,warning=TRUE,error=TRUE)
options(scipen=999) # removes sci notation
```

# Introduction

This package provides a generic implimentation of the iterative proportional 
fitting algorithm or [IPF](https://en.wikipedia.org/wiki/Iterative_proportional_fitting).

# First IPF Example

This section creates a random seed table and target values to illustrate 
how the package is used. The targets are specified for three separate clusters,
which could represent model TAZs, survey districts, or some other grouping of
geographic areas.

## Seed table creation
The seed table is the starting point for the IPF procedure.  In this example, 
we will use a made up summary table of survey data.  This fake table is a count 
of households by the number persons, workers and vehicles.  A sample of the 
table is shown for illustration.  The count is stored in the `weight` field.
```{r "seed creation", message=FALSE,warning=FALSE,error=FALSE}
library(dplyr)
library(tidyr)
library(ipfr)
library(htmlTable)

seed <- expand.grid(
  siz = c(1, 2, 3, 4),
  wrk = c(0, 1, 2, 3),
  veh = c(0, 1, 2, 3)
) %>% tbl_df()

set.seed(1)
seed$weight <- sample(1:10, nrow(seed), replace = TRUE)

seed %>%
  head() %>%
  htmlTable(
  align = "cccr", col.rgroup = c("none", "#F7F7F7"), rnames = FALSE
)
```


## Target creation
The number of households by size (e.g., 1-person, 2-person, etc.) is referred to
as a marginal distribution.  Often, from the Census, we know the total number of 
households by each individual marginal.  For example, we know the number of 
households by size:

  * 1 person: 150 households
  * 2 persons: 150 households
  * 3 persons: 100 households
  * 4+ persons: 100 households

This information becomes the target that the IPF process tries to match.

This information is created below for the distribution of households by size, 
number of workers, and number of vehicles.  The size table is shown as an example.
Across the top are the marginal categories (1-4). The cluster column represents
three separate geographies (counties, tracts, TAZs, etc.).
```{r "marginal creation"}

targets <- list()
targets$siz <- data_frame(
  cluster = c(1, 2, 3),
  `1` = c(100, 150, 0),
  `2` = c(100, 100, 0),
  `3` = c(150, 100, 0),
  `4` = c(150, 150, 0)
)
targets$wrk <- data_frame(
  cluster = c(1, 2, 3),
  `0` = c(100, 100, 0),
  `1` = c(150, 100, 0),
  `2` = c(150, 150, 0),
  `3` = c(100, 150, 0)
)
targets$veh <- data_frame(
  cluster = c(1, 2, 3),
  `0` = c(75, 175, 0),
  `1` = c(175, 75, 0),
  `2` = c(150, 150, 0),
  `3` = c(100, 100, 0)
)

targets$siz %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Size Targets"
  )  
```

## The Goal
We know the count of each marginal individually, but assume we don't have 
official information from the Census about the joint distribution.  What we want
to do is use the joint distribution from our survey, but ensure it agrees with 
the marginal information from the Census.

An initial test comparing sizes for cluster 1 shows they do not agree.
```{r "initial test"}
siz <- targets$siz %>%
  filter(cluster == 1) %>%
  gather(key = siz, value = target, -cluster) %>%
  mutate(siz = as.numeric(siz)) %>%
  select(-cluster)
  

seed %>%
  select(siz, weight) %>%
  group_by(siz) %>%
  summarize(seed = sum(weight)) %>%
  left_join(siz, by = "siz") %>%
  htmlTable(
    align = "crr", rnames = FALSE,
    caption = "Comparing size marginals for cluster 1"
  )
```

## Using IPF

We can use the ipfr package to change this.  For this example, the `verbose`
option is turned on to show what the tool reports upon completion. It reports 
the worst-performing marginal in the worst-performing cluster. If these
values are small, then you can be confident in the result.
```{r}
table <- ipf(seed, targets, verbose = TRUE)
```

A sample of the resulting table is shown below.
```{r}
table %>%
  mutate_all(funs(round(., 2))) %>%
  head() %>%
  htmlTable(
    align = "crr", rnames = FALSE
  )
```

## Summarizing Results

We can summarize the final table to see if it worked.
```{r}
table %>%
  select(-c(wrk, veh)) %>%
  group_by(cluster, siz) %>%
  summarize(total = round(sum(weight), 0)) %>%
  spread(key = siz, value = total) %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Size: Result"
  )

targets$siz %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Size: Target"
  )
```

Not only does the marginal distribution of size match, but also workers and vehicles.
```{r, echo=FALSE}
# Workers
table %>%
  select(-c(siz, veh)) %>%
  group_by(cluster, wrk) %>%
  summarize(total = round(sum(weight), 0)) %>%
  spread(key = wrk, value = total) %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Workers: Result"
  )

targets$wrk %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Workers: Target"
  )

# Vehciles
table %>%
  select(-c(siz, wrk)) %>%
  group_by(cluster, veh) %>%
  summarize(total = round(sum(weight), 0)) %>%
  spread(key = veh, value = total) %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Vehicles: Result"
  )

targets$veh %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Vehciles: Target"
  )
```

```{r}
# Saving seed and target for examples after example 2.
first_seed <- seed
first_targets <- targets
```


# Example 2: Combining Household and Person Marginals

In household survey expansion, it is common to want to control for certain
features that describe households, (like size), while controlling for other
attributes that describe people (like age). This is possible with `ipfr`.

First consider a household table of four households with varying sizes.

```{r}
hh_tbl <- data_frame(
  hhid = c(1, 2, 3, 4),
  siz = c(1, 2, 2, 1)
)

hh_tbl %>%
  htmlTable(
    rnames = FALSE, caption = "Survey Household Table"
  )
```

That household table has an associated person table shown below. `age_group`
takes on values between 1-4, and for clarity, let's say they represent the
following:

  * ag_1 Under 18
  * ag_2 Between 18 and 25
  * ag_3 Between 25 and 50
  * ag_4 Over 50

```{r}
per_tbl <- data_frame(
  hhid = c(1, 2, 2, 3, 3, 4),
  perno = c(1, 1, 2, 1, 2, 1),
  age_group = c("ag_4", "ag_3", "ag_1", "ag_4", "ag_4", "ag_2")
)

per_tbl %>%
  htmlTable(
    rnames = FALSE, caption = "Survey Person Table"
  )
```

A seed table with both attributes can be created. Note that each row is a unique
person, and that household attributes are repeated for households that have
more than one person.

```{r}
seed <- per_tbl %>%
  left_join(hh_tbl, by = "hhid") %>%
  mutate(weight = 1)

seed %>%
  htmlTable(
    rnames = FALSE, caption = "Combined Seed Table"
  )
```

Create targets similar to before (this time for one cluster to keep it simple).

```{r}
targets <- list()
targets$siz <- data_frame(
  cluster = c(1),
  `1` = c(50),
  `2` = c(150)
)
targets$age_group <- data_frame(
  cluster = c(1),
  ag_1 = 50,
  ag_2 = 50,
  ag_3 = 50,
  ag_4 = 50
)
```

Call the ipf function.

```{r}
result <- ipf(seed = seed, targets = targets)

result %>%
  mutate(weight = round(weight, 2)) %>%
  htmlTable(
    rnames = FALSE, caption = "IPF Result"
  )
```

The sum of the weight column is 200, which is the total number of households
in our target tables. The last step is simply to sum the weights by HHID to get
the final answer.

```{r}
result %>%
  group_by(hhid) %>%
  summarize(weight = round(sum(weight), 2)) %>%
  htmlTable(
    rnames = FALSE, caption = "Final Household Weights"
  )
```

Further checks on this simple data set show that both marginals were respected.

```{r}
result %>%
  group_by(age_group) %>%
  summarize(weight = round(sum(weight), 2)) %>%
  htmlTable(
    rnames = FALSE, caption = "Final Age Marginal Distribution"
  )

result %>%
  group_by(siz) %>%
  summarize(weight = round(sum(weight), 2)) %>%
  htmlTable(
    rnames = FALSE, caption = "Final Size Marginal Distribution"
  )
```

# Example 3: Survey Expansion using Geographical Clusters

In the first example, target values were specified for various clusters, but no
target information was appended to the seed table. In this case, `ipfr` assumes
that the full seed table is used as a starting point for every cluster.

In the case of survey expansion, it is common to define clusters in both the 
seed and target table. With this approach, only the seed records in a cluster
will be expanded to match that cluster's targets.

Accomplishing this is as simple as appending the `cluster` values to the seed 
table. We will create a new seed table, and append cluster values into a new
`cluster` field. Note that `cluster` is a special field name and must be used in
both seed and target tables for this to work.

```{r}
seed <- data_frame(
  hhid = c(1, 2, 3, 4),
  siz = c(1, 2, 1, 1),
  weight = c(1, 1, 1, 1),
  cluster = c(1, 1, 2, 2)
)

seed %>%
  htmlTable(
    rnames = FALSE, caption = "Seed Table with Cluster Appended"
  )
```

New targets are specified:

```{r}
targets <- list()
targets$siz <- data_frame(
  cluster = c(1, 2),
  `1` = c(75, 100),
  `2` = c(25, 150)
)
```

Run `ipf`. Note that it picks up a new error. When providing cluster values in the seed
table, `ipfr` checks to make sure that every marginal category appears in every cluster.
In this case, the full seed table does contain both size 1 and 2 records, but cluster
2 only has size 1.

```{r}
result <- ipf(seed = seed, targets = targets)
```

Fix the seed table and re-run `ipf`.

```{r}
seed <- data_frame(
  hhid = c(1, 2, 3, 4),
  siz = c(1, 2, 2, 1),
  weight = c(1, 1, 1, 1),
  cluster = c(1, 1, 2, 2)
)

result <- ipf(seed = seed, targets = targets)
```


In the resulting table, note that the first two rows in the seed table (in 
cluster 1) match the 100 household total of cluster 1, while the second two
rows match the 250 household total of cluster 2.

```{r}
result %>%
  htmlTable(
    rnames = FALSE, caption = "Survey Expansion to Clusters"
  )
```



# How IPFR addresses common IPF problems

This section will show how `ipfr` addresses some common problems found in basic
ipf procedures. It uses the example data from the first example.

```{r}
seed <- first_seed
targets <- first_targets
```


## Zero weights
IPF works by successively multiplying the table weights by factors.  Cells with 
a zero weight cannot be modified by this process.  As the number of zero weights 
increase, the flexibility of the process is reduced, and convergence becomes 
more difficult. `ipfr` solves this problem by setting a minimum weight for all 
cells to `.0001`.  This minimum weight can be adjusted using the `min_weight` 
parameter and should be arbitrarily small compared to your seed table weights.


## Missing seed information
Not every combination of marginal categories is required to be included in the 
seed table; however, at least one observation of each category must exist.  For example, the combination:

  * siz = 1
  * wrk = 1
  * veh = 0

may not have been observed in the survey, and thus may be missing from the seed 
table.  As long as other combinations of size-1 households exist (e.g. with 0 workers and 1 vehicle), `ipfr` will work fine.  On the other hand, if there are no observations of any size-1 households, `ipfr` will stop with an error message.

```{r, warning=TRUE}
missing_seed <- seed %>%
  filter(siz != 1)

error <- try(table <- ipf(missing_seed, targets))
error[1]
```

## Marginal agreement
In our example, the marginal distributions all add up to the same number for each zone (500, 500, and 0). If they disagreed on the total number of households, the standard IPF process could not converge.

Occasionally, marginal distributions might disagree slightly on the totals, particularly 
if they are derived from different sources.  When that happens, `ipfr` 
will still converge by matching the total of the first marginal table supplied 
while using the percentage distribution from the remaining tables.

```{r, warning=TRUE}
# Increase 1-person households from 100 to 500
targets$siz <- data_frame(
  cluster = c(1, 2, 3),
  `1` = c(500, 500, 0),
  `2` = c(100, 100, 0),
  `3` = c(150, 100, 0),
  `4` = c(150, 150, 0)
)

table <- ipf(seed, targets)
```

```{r, echo=FALSE}
# Size
table %>%
  select(-c(wrk, veh)) %>%
  group_by(cluster, siz) %>%
  summarize(total = round(sum(weight), 0)) %>%
  spread(key = siz, value = total) %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Size: Result"
  )

targets$siz %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Size: Target"
  )

# Workers
table %>%
  select(-c(siz, veh)) %>%
  group_by(cluster, wrk) %>%
  summarize(total = round(sum(weight), 0)) %>%
  spread(key = wrk, value = total) %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Workers: Result"
  )

targets$wrk %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Workers: Target"
  )

# Vehciles
table %>%
  select(-c(siz, wrk)) %>%
  group_by(cluster, veh) %>%
  summarize(total = round(sum(weight), 0)) %>%
  spread(key = veh, value = total) %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Vehicles: Result"
  )

targets$veh %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Vehciles: Target"
  )
```

## Partial marginals
In some applications of IPF, there may not be a need to control every category of every marginal.  In this example, perhaps you want to control 1-, 2-, and 3-person households, but are not concerned with 4+ person households. `ipfr` can handle a partial marginal table.

```{r}
partial_targets <- targets
partial_targets$siz <- data_frame(
  cluster = c(1, 2, 3),
  `1` = c(100, 150, 0),
  `2` = c(100, 100, 0),
  `3` = c(150, 100, 0)
)

table <- ipf(seed, partial_targets)
```

As shown below, the marginal distributions of size categories 1, 2, and 3 are fairly close to their target values, while 4+ is not.

```{r}
table %>%
  select(-c(wrk, veh)) %>%
  group_by(cluster, siz) %>%
  summarize(total = round(sum(weight), 0)) %>%
  spread(key = siz, value = total) %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Size: Result"
  )

partial_targets$siz %>%
  htmlTable(
    align = "crr", rnames = FALSE, caption = "Size: Target"
  )
```